Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance

### 2017-08-08 ms_pre_post.Rmd
Minor changes

In the context of inbreeding as the postulated driver of reproductive strategy evolution, the absolute and relative frequencies of alleles underlying pre-copulatory and post-copulatory inbreeding avoidance and polyandry will be affected not only by the magnitudes of positive indirect selection stemming from reduced inbreeding depression in females' offspring, but also by the magnitudes of direct negative selection on resulting phenotypes (i.e., the direct fitness costs of expressing each reproductive strategy). Empirical studies have demonstrated diverse costs of mating and mate choice, for example, including energetic costs of developing, maintaining, or enacting necessary physiologies [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]; increased risks of predation or disease stemming from increased mate-searching or mating [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]; increased risk of complete mating or fertilisation failure given extreme choosiness [@Kokko2013]; and risks of harm stemming from sexual conflict over fertilisation [e.g., @Rowe1994]. If the relative costs of pre-copulatory and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly strategy might become fixed over generations, while alleles underlying the more costly strategy might go extinct, especially if their effects become redundant following evolution of the less costly strategy. Evolutionary dynamics will therefore depend on the inbreeding avoidance benefit of each strategy, which might also differ and vary in relation to the evolution of the other strategy. Dynamic models that track the frequencies of alleles underlying multiple potentially costly inbreeding avoidance strategies that might influence each other's evolution across generations, and that are enacted among relatives that themselves result from reproductive strategies and inbreeding depression expressed in previous generations, are consequently useful to understand and predict evolutionary outcomes.

We use individual-based modelling to address three key questions regarding evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given opportunity for inbreeding and strong inbreeding depression. First, does evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance facilitate evolution of polyandry that is otherwise severely constrained by increasing relatedness of polyandrous females to their additional mates? Second, how do costs associated with polyandry and pre-copulatory and post-copulatory inbreeding avoidance affect evolutionary outcomes and, in particular, the long-term persistence of these reproductive strategies given cost asymmetry? Third, how is selection for initially rare pre-copulatory or post-copulatory inbreeding avoidance affected if the other strategy of inbreeding avoidance is already fixed? To address each question, we isolate the effect of the biological mechanism of interest (i.e., post-copulatory inbreeding avoidance, cost asymmetry, and strategy pre-existence, respectively) by comparing simulations with the mechanism present versus absent with otherwise identical parameter values and conditions. We thereby illustrate how the simultaneous evolution of multiple interacting degenerate reproductive strategies can generate diverse evolutionary outcomes.

@@

Phenotypic values of the three reproductive strategy traits each incur set costs that combine independently to increase the probability that a focal female will die before mating. Numerous forms and mechanisms of direct costs of reproductive strategies could potentially be hypothesised and modelled; the most appropriate formulation depends on the question posed (see Discussion). By allowing costs of the three strategies to be directly and independently controlled, our model facilitates direct comparison of evolution of pre-copulatory versus post-copulatory inbreeding avoidance given known cost asymmetries. Such costs on female survival probability are biologically reasonable. For example, polyandrous females that undertake increased mate searching or mating can experience increased predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice can increase the risk of mortality due to harm caused by sexual conflict over mating, and also risk complete mating failure [which equates to pre-reproductive mortality in semelparous organisms; @Rowe1994; @Kokko2013]. Finally, females that express post-copulatory choice can pay up-front energetic costs, which might result in trade-offs with survival due to developing physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014].

@@

To address how asymmetric costs associated with pre-copulatory and post-copulatory inbreeding avoidance and polyandry affect the long-term persistence of reproductive strategies, we quantify the change in $M_{a}$ and $F_{a}$ over generations in simulations where pre-copulatory inbreeding strategy was cost free ($c_{M}=0$) but post-copulatory inbreeding strategy was moderately costly ($c_{F}=0.02$), and vice versa. We compare evolutionary trajectories with those of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). Previous modelling using similar genetic architecture suggests that a cost of $0.02$ imposes strong but not overwhelming direct negative selection on polyandry [@Duthie]. This value is therefore appropriate to illustrate the different evolutionary consequences that could result from asymmetrical costs (see Supporting Information for additional cost value combinations).

@@

When post-copulatory inbreeding avoidance alleles ($F_{a}$) had no effect (i.e., $F_{g}$ values were fixed to zero), meaning that $F_{p}$ could not evolve, $P_{a}$ alleles underlying polyandry decreased to negative values over generations (red lines, Figure 1A,C,E,G). This shows that despite strong inbreeding depression in offspring viability, there is selection against unconditional polyandry even given zero direct cost ($c_{P}=0$; Figure 1A). This is because $M_{a}$ values became negative over generations, meaning that females typically avoided inbreeding through their initial mating (blue lines, Figure 1A,C,E,G). Polyandrous females that subsequently sampled more males from the available population were consequently more likely to mate with some relatives and hence produce some inbred offspring with low viability [@Duthie].

@@

Evolution of different reproductive strategies cannot be presumed to occur in isolation from one another. Rather, there is likely to be considerable potential for feedbacks and degeneracy (i.e., functional redundancy) among interacting phenotypes. For example, inbreeding avoidance could be manifested through both pre-copulatory and post-copulatory mechanisms and associated polyandry, meaning that simultaneous evolution of each phenotype might be affected by degeneracy, in addition to trait-specific benefits and costs.

We used individual-based modelling to highlight fundamental but theoretically under-developed relationships between evolution of polyandry and pre-copulatory versus post-copulatory inbreeding strategy given (1) hard constraints on evolution of post-copulatory inbreeding strategy, (2) asymmetric costs of pre-copulatory and post-copulatory inbreeding strategy, and (3) evolution of one inbreeding strategy phenotype given pre-existence of the other. We present our model and simulation results as tools for thinking more clearly about the dynamics of simultaneously evolving reproductive strategies, not for making accurate quantitative predictions for any specific system. While the logic of our model can be usefully applied to construct general hypotheses within and across empirical systems, accurate quantitative prediction would require additional empirical detail and data for model parameterisation.

@@

Model assumptions and extensions
-------------------------------------------------------------------

To provide a general conceptual comparison of the effects of direct costs across phenotypes, we modelled all costs as analogous increased probabilities of total reproductive failure. This cost formulation reflects empirical observations in some populations [see Model; e.g., @Rowe1994; @Koga1998; @Gasparini2011], and is therefore a biologically realistic means of standardising costs across traits. However, different forms of costs could be incorporated into future models designed to predict evolutionary dynamics that might occur in specific systems. Models could then be further developed such that costs arise from explicit reproductive mechanisms, requiring further biological detail.

For example, @Pomiankowski1987 identified four cost categories relevant to mating frequency and mate choice, including elevated risks of predation or disease transmission, and time or energy expenditure. Polyandrous females might experience increased risk of disease transmission [@Roberts2015], a cost that would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating [e.g., @Arnqvist2005; @Parker2006]. Inbreeding theory predicts that males should be more tolerant of inbreeding than females, leading to sexual conflict over inbreeding in mating encounters [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Future models could therefore explicitly consider sexual conflict over both polyandry and pre-copulatory inbreeding, and hence capture internally-consistent mechanistic costs. Further, sexual conflict might also affect post-copulatory inbreeding. For example, when female guppies (*Poecilia reticulata*) were artificially inseminated with equal quantitites of sperm from full-siblings and unrelated males, more eggs were fertilised by unrelated males because the velocities of full sibling's sperm were reduced by females' ovarian fluids [@Gasparini2011]. In black field crickets (*Teleogryllus commodus*), females attempt to remove the spermatophores of unwanted males after copulation, and are capable of controlling sperm transfer to spermatheca after copulation occurs [@Bussiere2006; @Tuni2013]. Model extensions might therefore incorporate such post-copulatory sexual conflict to better understand female and male reproductive strategy evolution.

We restricted our model to simulate how evolution of traits was affected by direct costs and indirect benefits due to inbreeding avoidance. Although natural populations will include additional beneficial traits that might affect evolution of mating strategies, targeting our focus on traits affecting inbreeding and polyandry allows us to better isolate their effects and therefore better develop general hypotheses. Future models could be further developed to include additional benefits and costs of reproductive strategies, and benefit and cost values could be parameterised using data from empirical populations, hence enabling more quantitative predictions for specific systems. The addition of such mechanistic detail, and the inclusion of data from populations of interest to parameterise model extensions, could ultimately improve understanding and prediction of reproductive strategy evolution and thereby link theory, modelling, and empirical hypothesis testing.

>>>>

In the context of inbreeding avoidance as a key hypothesised driver of reproductive strategy evolution, the absolute and relative frequencies of alleles underlying pre-copulatory and post-copulatory inbreeding avoidance and polyandry will be affected not only by the magnitudes of positive indirect selection stemming from reduced inbreeding depression in females' offspring, but also by the magnitudes of direct negative selection on resulting phenotypes (i.e., the direct fitness costs of expressing each reproductive strategy). Empirical studies have demonstrated diverse costs of mating and mate choice, for example, including energetic costs of developing, maintaining, or enacting necessary physiologies [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014b]; increased risks of predation or disease stemming from increased mate-searching or mating [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]; increased risk of complete mating or fertilisation failure given extreme choosiness [@Kokko2013]; and risks of harm stemming from sexual conflict over fertilisation [e.g., @Rowe1994]. If the relative costs of pre-copulatory and post-copulatory inbreeding avoidance differ, then alleles underlying the less costly strategy might become fixed over generations, while alleles underlying the more costly strategy might go extinct, especially if their effects become redundant following evolution of the less costly strategy. Evolutionary dynamics will therefore depend on the inbreeding avoidance benefit of each strategy, which might also differ and vary in relation to the evolution of the other strategy. Dynamic models that track the frequencies of alleles underlying multiple potentially costly inbreeding avoidance strategies that might influence each other's evolution across generations, and that are enacted among relatives that themselves result from reproductive strategies and inbreeding depression expressed in previous generations, are consequently useful to understand and predict evolutionary outcomes.

We use individual-based modelling to address three key questions regarding evolution of pre-copulatory and post-copulatory inbreeding avoidance and associated polyandry given opportunity for inbreeding and strong inbreeding depression. First, does evolution of post-copulatory inbreeding avoidance alongside pre-copulatory inbreeding avoidance facilitate evolution of polyandry that is otherwise severely constrained by increasing relatedness of polyandrous females to their additional mates? Second, how do costs associated with polyandry and pre-copulatory and post-copulatory inbreeding avoidance affect evolutionary outcomes and, in particular, the long-term persistence of these reproductive strategies given cost asymmetry? Third, how is selection for initially rare pre-copulatory or post-copulatory inbreeding avoidance affected if the other strategy of inbreeding avoidance is already fixed? To address these questions, our model is designed to allow each biological mechanism of interest (i.e., post-copulatory inbreeding avoidance, cost asymmetry, and strategy pre-existence, respectively) to be isolated, thereby facilitating direct comparison of simulations with the mechanism present versus absent with otherwise identical parameter values and conditions. We thereby illustrate how the simultaneous evolution of multiple interacting degenerate reproductive strategies can generate diverse evolutionary outcomes.

@@

Phenotypic values of the three reproductive strategy traits each incur set costs that combine to independently increase the probability that a focal female will die before mating. Numerous forms and mechanisms of direct costs of reproductive strategies could potentially be hypothesised and modelled; the most appropriate formulation depends on the question posed (see Discussion). By allowing costs of the three strategies to be directly and independently controlled, our model facilitates direct comparison of evolution of pre-copulatory versus post-copulatory inbreeding avoidance given known cost asymmetries. Such costs on female survival probability are biologically reasonable. For example, polyandrous females that undertake increased mate searching or mating can experience increased predation risk [e.g., @Rowe1988; @Ronkainen1994; @Koga1998]. Females that express pre-copulatory choice can increase the risk of mortality due to harm caused by sexual conflict over mating, and also risk complete mating failure [which equates to pre-reproductive mortality in semelparous organisms; @Rowe1994; @Kokko2013]. Finally, females that express post-copulatory choice can pay up-front energetic costs, which might result in trade-offs with survival due to developing physiological or biochemical mechanisms needed to store sperm and successfully bias fertilisation [e.g., @Gasparini2011; @Tuni2013; @Fitzpatrick2014].

@@

To address how asymmetric costs associated with pre-copulatory and post-copulatory inbreeding avoidance and polyandry affect the long-term persistence of reproductive strategies, we quantify the change in $M_{a}$ and $F_{a}$ over generations in simulations where pre-copulatory inbreeding strategy was cost free ($c_{M}=0$) but post-copulatory inbreeding strategy was moderately costly ($c_{F}=0.02$), and vice versa. We compare evolutionary trajectories with those of a costly strategy in the absence of evolution of an alternative strategy (e.g., evolution of pre-copulatory inbreeding strategy when post-copulatory inbreeding strategy phenotype is fixed at zero, $F_{p}=0$). Previous modelling using similar genetic architecture suggests that a cost of $0.02$ imposes strong but not overwhelming direct negative selection on polyandry [@Duthie]. This value is therefore appropriate to illustrate the different evolutionary consequences that could result from asymmetrical costs, but simulations with additional cost value combinations are provided in Supporting Information.

@@

When post-copulatory inbreeding avoidance alleles ($F_{a}$) had no effect (i.e., $F_{g}$ values were fixed to zero), meaning that $F_{p}$ could not evolve, $P_{a}$ alleles underlying polyandry decreased to negative values over generations (red lines, Figure 1A,C,E,G). This shows that despite strong inbreeding depression in offspring viability, there is selection against unconditional polyandry even given zero direct cost ($c_{P}=0$; Figure 1A). This is because $M_{a}$ values became negative over generations, meaning that females typically avoided inbreeding through their initial mating (blue lines, Figure 1A,C,E,G). Polyandrous females that subsequently sampled more males from the available population were consequently more likely to mate with some relatives and hence produce some inbred offspring with low viability [see also @Duthie].

@@

Different reproductive strategies cannot be presumed to evolve in isolation from one another. Rather, there is likely to be considerable potential for feedbacks and degeneracy (i.e., functional redundancy) among interacting phenotypes. For example, inbreeding avoidance could be manifested through both pre-copulatory and post-copulatory mechanisms and associated polyandry, meaning that simultaneous evolution of each phenotype might be affected by degeneracy, in addition to trait-specific benefits and costs.

We used individual-based modelling to highlight fundamental but theoretically under-developed relationships between evolution of polyandry and pre-copulatory versus post-copulatory inbreeding strategy given (1) hard constraints on evolution of post-copulatory inbreeding strategy, (2) asymmetric costs of pre-copulatory and post-copulatory inbreeding strategy, and (3) evolution of one inbreeding strategy phenotype given pre-existence of the other. Our current model and simulation results thereby provide tools for thinking more clearly about the dynamics of simultaneously evolving reproductive strategies in the context of polyandry and inbreeding avoidance.

@@

Model structure, assumptions, and extensions
-------------------------------------------------------------------

While the logic of our current model can be usefully applied to construct general hypotheses within and across empirical systems, accurate quantitative prediction for specific systems would require additional empirical detail and data for model parameterisation. To facilitate general conceptual comparison of the effects of direct costs across phenotypes, we modelled all costs as analogous increased probabilities of female mortality and hence total reproductive failure. This cost formulation reflects empirical observations in some populations [see Model; e.g., @Rowe1994; @Koga1998; @Gasparini2011], and is therefore a biologically realistic method of standardising costs across traits. However, different forms of costs could be incorporated into future models designed to predict specific evolutionary dynamics. Models could then be further developed such that costs arise from explicit reproductive mechanisms, requiring further biological detail.

For example, @Pomiankowski1987 identified four cost categories relevant to mating frequency and mate choice, including elevated risks of predation or disease transmission, and time or energy expenditure. Polyandrous females might experience increased risk of disease transmission [@Roberts2015], a cost that would more realistically apply to a female's realised number of mates rather than her tendency for polyandry. Polyandrous females might also risk harm caused by sexual conflict over multiple mating [e.g., @Arnqvist2005; @Parker2006]. Inbreeding theory predicts that males should be more tolerant of inbreeding than females, leading to sexual conflict over inbreeding in mating encounters [@Parker1979; @Parker2006; @Kokko2006; @Duthie2015a]. Future models could therefore explicitly consider sexual conflict over both polyandry and pre-copulatory inbreeding, and hence capture internally-consistent mechanistic costs. Further, sexual conflict might also affect post-copulatory inbreeding. For example, when female guppies (*Poecilia reticulata*) were artificially inseminated with equal quantitites of sperm from full-siblings and unrelated males, more eggs were fertilised by unrelated males because the velocities of full sibling's sperm were reduced by females' ovarian fluids [@Gasparini2011]. In black field crickets (*Teleogryllus commodus*), females attempt to remove the spermatophores of unwanted males after copulation, and are capable of controlling sperm transfer to spermatheca after copulation occurs [@Bussiere2006; @Tuni2013]. Future models could therefore explicitly incorporate such mechanisms in order to better understand effects of post-copulatory sexual conflict on female and male reproductive strategy evolution.

We restricted our current model to examine how reproductive strategy evolution was affected by direct costs and indirect benefits stemming from inbreeding avoidance, thereby isolating such effects and explicitly addressing key general hypotheses regarding the effects of inbreeding depression. In natural populations, reproductive strategy evolution will also be affected by other costs and benefits, for example stemming from additive genetic effects (i.e., pre-copulatory or post-copulatory mate choice for 'good genes'). Future modelling could consequently include multiple benefits and costs of reproductive strategies, which could then be parameterised using empirical data. The addition of such mechanistic detail, and the inclusion of data from populations of interest to parameterise model extensions, could ultimately improve understanding and prediction of reproductive strategy evolution and thereby link theory, modelling, and empirical hypothesis testing.
